---
title: "Assignment 8 - Plotting amounts, proportions, and using color"
author: "Luke Henslee"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r, include=FALSE, results=FALSE}
library(tidyverse)
library(ggplot2)
```

### Task 1: Heat maps  

#### 1) Import data and subset 1950 -- 2017 observations   

```{r}
npgo.pdo <- read.csv("npgo-pdo.csv")

npgo.pdo.sub <- npgo.pdo %>% 
  filter(npgo.pdo$Year < 2018)

# Turn months into factors and set levels to the R constant 'month.abb'
npgo.pdo.sub$Mon <- factor(npgo.pdo.sub$Mon, levels = month.abb)
```
<br/><br/>

#### 2) NPGO by month and year  

To visualize fluxuations in NPGO by month and year, we will create a heatmap which
uses the 'viridis' color palette, which is common for visualizing continuous 
variables. 

```{r}
ggplot(data = npgo.pdo.sub, aes(x = Year, y = Mon)) +
  geom_raster(aes(fill = NPGO)) +
  scale_y_discrete(limits = sort(npgo.pdo.sub$Mon, decreasing = TRUE)) +
  scale_x_continuous(breaks = c(1950, 1960, 1970, 1980, 1990, 2000, 2010, 2017)) +
  scale_fill_viridis_c(name = NULL) +
  theme_minimal() +
  theme(panel.grid.major = element_blank()) +
  labs(title = "NPGO anomalies by month and year", x = "Year", y = "Month") +
  geom_vline(xintercept = 1977, color = "white", linetype = "dashed", size = 1.5) +
  geom_vline(xintercept = 1989, color = "white", linetype = "dashed", size = 1.5) 
```
<br/><br/>

#### 3) PDO by month and year

This time we are going to try an unconventional color pallete from the 
'wesanderson' package.

```{r}
library(wesanderson)

ggplot(data = npgo.pdo.sub, aes(x = Year, y = Mon)) +
  geom_raster(aes(fill = PDO)) +
  scale_y_discrete(limits = sort(npgo.pdo.sub$Mon, decreasing = TRUE)) +
  scale_x_continuous(breaks = c(1950, 1960, 1970, 1980, 1990, 2000, 2010, 2017)) +
  scale_fill_gradientn(colors = wes_palette("Zissou1", 100, type = "continuous"), name = NULL)+
  theme_minimal() +
  theme(panel.grid.major = element_blank()) +
  labs(title = "PDO anomalies by month and year", x = "Year", y = "Month")+
  geom_vline(xintercept = 1977, color = "white", linetype = "dashed", size = 1.5) +
  geom_vline(xintercept = 1989, color = "white", linetype = "dashed", size = 1.5)
```

It's a little headachey, but kind of interesting.... It actually does a pretty good job of showing the values near the mean and those that are extreme anamolies in either direction.   
<br/><br/>

#### 4) Visually depict a relationship between NPGO and PDO by month and year    
To show a possible relationship between NPGO and PDO I'm going to plot the absolute values of the differences between the two by month and year. Cells which have values near zero indicate that PDO and NPGO fluctuated together. A plot of mostly the same color should indicate a strong relationship between NPGO and PDO. 

```{r}
# Create new variable 'abs.diff'
npgo.pdo.sub$abs.diff <- abs(npgo.pdo.sub$NPGO - npgo.pdo.sub$PDO)

ggplot(data = npgo.pdo.sub, aes(x = Year, y = Mon)) +
  geom_raster(aes(fill = abs.diff)) +
  scale_y_discrete(limits = sort(npgo.pdo.sub$Mon, decreasing = TRUE)) +
  scale_x_continuous(breaks = c(1950, 1960, 1970, 1980, 1990, 2000, 2010, 2017)) +
  scale_fill_viridis_c(name = NULL) +
  theme_minimal() +
  theme(panel.grid.major = element_blank()) +
  labs(title = "Difference between NPGO and PGO anomalies by month and year", x = "Year", y = "Month") +
  geom_vline(xintercept = 1977, color = "white", linetype = "dashed", size = 1.5) +
  geom_vline(xintercept = 1989, color = "white", linetype = "dashed", size = 1.5) 
```

Okay, not bad. I'm going to simplify things by assigning a '0' to negative anomalies and '1' to positive anomolies. Then I will plot the difference between NPGO and PDO. Months which have the same directionality of their anomalies will add up to zero and months with different directionality will add up to one. There will only be two colors in this heat map.

```{r}
dir.npgo <- vector(length = nrow(npgo.pdo.sub))

for(i in 1:length(dir)){
  if(npgo.pdo.sub[i,3] > 0) {
    dir.npgo[i] = 1
  } else {
    dir.npgo[i] = 0
  }
}

dir.pdo <- vector(length = nrow(npgo.pdo.sub))

for(i in 1:length(dir)){
  if(npgo.pdo.sub[i,4] > 0) {
    dir.pdo[i] = 1
  } else {
    dir.pdo[i] = 0
  }
}

npgo.pdo.sub$dir <- factor(abs(dir.npgo - dir.pdo))
```

Okay, now let's make the plot. 

```{r}
ggplot(data = npgo.pdo.sub, aes(x = Year, y = Mon)) +
  geom_raster(aes(fill = dir)) +
  scale_y_discrete(limits = sort(npgo.pdo.sub$Mon, decreasing = TRUE)) +
  scale_x_continuous(breaks = c(1950, 1960, 1970, 1980, 1990, 2000, 2010, 2017)) +
  scale_fill_grey(start = 0.9, end = 0.2, name = NULL, labels = c("Trend together", "Trend apart")) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  labs(title = "Anomaly trends for the NPGO and PGO anomalies by month and year", x = "Year", y = "Month") +
  geom_vline(xintercept = 1977, color = "red", linetype = "dashed", size = 1.5) +
  geom_vline(xintercept = 1989, color = "red", linetype = "dashed", size = 1.5) 
```

<br/><br/>

I kind of like it! I don't know if it qualifies as a heatmap, but it does show us some interesting visuals about the relationship between the two indicies.

#### 5) What do you notice in each of the plots near the regime shifts at 1977 and 1989?  
So we've gone back through and added dotted vertical lines marking the regime shifts to our plots. The heatmap showing NPGO anomolies indicates that around 1977, differences from the mean began trending a little more toward the positive, and a lot more positive after 1989 and more extreme. Almost all anomalies in the PDO index between the regime shifts are in the positive direction, and after 1989 they get more extreme and mostly stay positive.  It looks like the NPGO and PDO were more likely to trend together and in the same direction before the 1977 regime shift, and began to fluctuate in different directions more often afterward. From 1989 on, anomalies in the indicies have been more extreme and tend to deviate from the mean in different directions. 

### Task 2: Plotting multiple proportions
#### 1) Import data

```{r}
narw <- read.csv("NARW_mort.csv", fileEncoding="UTF-8-BOM")
```
<br/><br/>

#### 2) Create mosiac plot

```{r}
library(ggmosaic)

ggplot(data = narw) +
  geom_mosaic(aes(x = product(Country, CauseDeath), fill = Sex)) +
  scale_fill_manual(values = c("#66CC00", "#0080FF"), labels = c("Female", "Male")) +
  annotate("text", x = .15, y = .5, label = "Entanglement", angle = 90, size = 10, alpha = .5) +
    annotate("text", x = .48, y = .5, label = "Undetermined", angle = 90, size = 10, alpha = .5) +
  annotate("text", x = .85, y = .5, label = "Vessel Strike", angle = 90, size = 10, alpha = .5) +
  labs(y = "Country", x = "Cause of Death", title = "Right Whale mortality by sex, country, and cause of death") +
  theme_minimal() +
  theme(panel.grid.major = element_blank()) +
  theme(axis.text.x = element_blank(), axis.ticks = element_blank())
```


```{r}
ggplot(narw, aes(x = Country, y = CauseDeath, fill = Sex)) +
  geom_point(shape=21, position = position_jitter(w = .2, h = .2), size = 4, alpha = .5) +
  scale_fill_manual(values = c("#66CC00", "#0080FF")) +
  facet_grid(CauseDeath~Year) +
  labs(y = "Cause of death", x = "Country", title = "Right Whale mortality by sex, country, and cause of death") +
  theme_bw() +
  theme(axis.text.y = element_blank(), axis.ticks = element_blank())
```





